Hierarchical time series forecasting in Emergency Medical Services

### 2022-11-16 manuscript/01-introduction.qmd
Spelling

In this paper, we address this gap by investigating the application of  hierarchical forecasting approaches in the EMS using daily time series of verified incidents from 2015 to 2020 in a major ambulance service in Great Britain. The data has hierarchical and grouped structures with hierarchies at the national, control (i.e. sub-national), health board (i.e. regional) and groups by priority and nature of incidents. We produce not only the point forecast but also the forecast distribution across all levels, which is critical for an effective planning and associated risk management. We compare the point and probabilistic forecast accuracy of the independent forecasts, bottom-up and optimal reconciliation approaches. We first generate independent/based forecasts using Exponential Smoothing State Space (ETS),  Generalised Linear Model (GLM), Poisson regression, a simple empirical distribution and an ensemble method followed by applying bottom-up and optimal reconciliation approaches. Forecast performance is assessed by the Root Mean Squared Scaled Error (RMSSE) for point forecasts and the Continuous Ranked Probability Score (CRPS) for the probabilistic forecasts. 
This paper complies with the principles of the reproducibility  [@stodden2013best; @boylan2015reproducibility]. Therefore, the study could equally be applied to any healthcare service (e.g. emergency department, primary or social care) subject to the time series having a hierarchical and/or grouped structure, which is generally the case in the healthcare sector.

>>>>

In this paper, we address this gap by investigating the application of  hierarchical forecasting approaches in the EMS using daily time series of verified incidents from 2015 to 2020 in a major ambulance service in Great Britain. The data has hierarchical and grouped structures with hierarchies at the national, control (i.e. sub-national), health board (i.e. regional) and groups by priority and nature of incidents. We produce not only the point forecast but also the forecast distribution across all levels, which is critical for an effective planning and associated risk management. We compare the point and probabilistic forecast accuracy of the independent forecasts, bottom-up and optimal reconciliation approaches. We first generate independent/based forecasts using Exponential Smoothing State Space (ETS),  Generalized Linear Model (GLM), Poisson regression, a simple empirical distribution and an ensemble method followed by applying bottom-up and optimal reconciliation approaches. Forecast performance is assessed by the Root Mean Squared Scaled Error (RMSSE) for point forecasts and the Continuous Ranked Probability Score (CRPS) for the probabilistic forecasts.
This paper complies with the principles of the reproducibility  [@stodden2013best; @boylan2015reproducibility]. Therefore, the study could equally be applied to any healthcare service (e.g. emergency department, primary or social care) subject to the time series having a hierarchical and/or grouped structure, which is generally the case in the healthcare sector.

### 2022-11-16 manuscript/02-literature.qmd
Spelling

There are generally two main streams of research related to forecasting ambulance demand in EMS: i) the first stream focuses on the application of time series methods and regression approaches on forecasting aggregate ambulance demand [@vile2012predicting; @sasaki2010using]; ii) the second stream considers forecasting EMS demand in a more finer temporal and geographical granularitties by employing temporal-spatial prediction methods [@zhou2016predicting;@zhou2016predictinglit]. The focus of our study is related to the first stream of research.

>>>>

There are generally two main streams of research related to forecasting ambulance demand in EMS: i) the first stream focuses on the application of time series methods and regression approaches on forecasting aggregate ambulance demand [@vile2012predicting; @sasaki2010using]; ii) the second stream considers forecasting EMS demand in a more finer temporal and geographical granularities by employing temporal-spatial prediction methods [@zhou2016predicting;@zhou2016predictinglit]. The focus of our study is related to the first stream of research.

### 2022-11-16 manuscript/03-experiment.qmd
Spelling

We are interested in generating forecasts to inform the planning horizon of $ph= 42$ days, interested by planners in the ambulance services trust. The forecast horizon in this study is $fh = 2 \times ph$ days ahead (2*42 days planning horizon). This is because the planning is generally freezed for $ph$ days and considering a forecast horizon of $ph$ days might not be helpful for planning. While forecasts are generated for $2 \times ph$ days ahead, performance evaluation is only assessed based on the last $ph$ days and not the $2 \times ph$ days. The forecasts are produced for the holdout of 365 days using time series cross-validation [@hyndman2021forecasting]. 
In the following section, we discuss the dataset, describe the forecasting methods used to generate base forecasts and present the point and probabilistic accuracy measures.

@@

```{r}
#| label: fig-hierarchy
#| out.width: "60%"
#| fig-cap: "The hierarchical and grouped structure of attended inciddents (ambulance demand)"
#| fig-subcap: 
#|   - "Hierarchical structure: Attended inciddents in all country is disaggregated into 3 control areas /regions and then into 7 different healrhboard at the bottom"
#|   - "Grouped structure:  Incidents could be grouped into priority (i.e. Red, Amber & Green) and the nature of attenedd incident (i.e. there are 35 different nature of incidents including chest pain , breahtign problems, heart attack, stroke, and son on). The symbol * refers to the crossed attributes between hierarchical and grouped levels."
#| layout: [[50],[-2], [38]]

>>>>

We are interested in generating forecasts to inform the planning horizon of $ph= 42$ days, interested by planners in the ambulance services trust. The forecast horizon in this study is $fh = 2 \times ph$ days ahead (2*42 days planning horizon). This is because the planning is generally frozen for $ph$ days and considering a forecast horizon of $ph$ days might not be helpful for planning. While forecasts are generated for $2 \times ph$ days ahead, performance evaluation is only assessed based on the last $ph$ days and not the $2 \times ph$ days. The forecasts are produced for the holdout of 365 days using time series cross-validation [@hyndman2021forecasting].
In the following section, we discuss the dataset, describe the forecasting methods used to generate base forecasts and present the point and probabilistic accuracy measures.

@@

```{r}
#| label: fig-hierarchy
#| out.width: "60%"
#| fig-cap: "The hierarchical and grouped structure of attended incidents (ambulance demand)"
#| fig-subcap:
#|   - "Hierarchical structure: Attended incidents in all country is disaggregated into 3 control areas /regions and then into 7 different healthboard at the bottom"
#|   - "Grouped structure:  Incidents could be grouped into priority (i.e. Red, Amber & Green) and the nature of attended incident (i.e. there are 35 different nature of incidents including chest pain , breathing problems, heart attack, stroke, and so on). The symbol * refers to the crossed attributes between hierarchical and grouped levels."
#| layout: [[50],[-2], [38]]
